Supplementary Information

### 2020-08-05 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

*other contributers may move up in the list if they contribute a lot*

@@

Once $DBH$ had been reconstructed, we calculated aboveground biomass growth increments ($\Delta AGB$) based on $DBH$ and $\Delta r$.
Specifically, for each year $y$, we used biomass allometries to estimate $ABG_y$ based on $DBH_y$ and $AGB_{y+1}$ based on [$DBH_y + \Delta r$]. 
We then calculated $\Delta AGB_y$ as [$AGB_{y+1}-AGB_y$]. 
For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
For BCI, biomass allometries were sourced from the BIOMASS package (REF).

@@

First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET). 
These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.

@@

Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
If PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
If it came out as the best variable in one but not both of these groups, it was dropped from the analysis, on the logic that it was an inferior predictor to a related variable.
We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.

A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)

@@

Second, we combined DBH, climate, and atmospheric pollution data in a GLS model (Fig. 1).
[**DESCRIBE GLS**]

@@

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if identified as the top variable for both the temperature and cloudiness groups. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)

@@

**Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS). 
This contrasts with patterns observed at the stand level [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back; [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.
Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al. 2006; check Piponiot et al. in prep). [EXPAND]
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

>>>>

*other contributors may move up in the list if they contribute a lot*

@@

Once $DBH$ had been reconstructed, we calculated aboveground biomass growth increments ($\Delta AGB$) based on $DBH$ and $\Delta r$.
Specifically, for each year $y$, we used biomass allometries to estimate $ABG_y$ based on $DBH_y$ and $AGB_{y+1}$ based on [$DBH_y + \Delta r$]. 
We then calculated $\Delta AGB_y$ as [$AGB_{y+1}-AGB_y$]. 
For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).

@@

First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to cloud cover and temperature variability ("cloud" group; daily temperature range, percent cloud cover, and again PET). 
Because the variables in the cloud group are linked to both temperature and precipitation, this group was given lowest precedence in the variable selection process.
These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.

@@

Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.
We checked for collinearity among the selected climate variables, removing cloud group variables when they were *too closely correlated* with temperature or precipitation group variables. 
Furthermore, we removed any variable with a variance inflation factor > 10. (**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))

A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek (*and NM*?). 
There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most significant climate drivers.
(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)

@@

Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
We checked for collinearity among the full set of variables and removed any variable with a variance inflation factor > 10. (**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))
For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree. We  also ran models with a spline of DBH (3 knots).
Across all possible models, variable AICc weights were summed to determine which drivers are most important. We considered the model containing those variables as the top model.

@@

![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if included in the model. For each species, relationships are plotted if included in top model. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)

@@

@@

**Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS). 
This contrasts with patterns observed at the stand level [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al. in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back; [Clark et al. 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al. 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
This also contrasts with in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), where we show a positive relationship of radial growth to DBH from tree-rings.
*Helene Muller-Landau argues that, from tree scaling/geometry perspective, $\Delta r$ is the closest to being size-independent (as opposed to BAI) (see [Francis et al. (2017)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12604)).* 
Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.
Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al. 2006; check Piponiot et al. in prep). [EXPAND]
$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.
